---
title: "Final paper 2025"
author: "Aude Dejonghe"
format: 
  pdf: default
  html: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
#| message: false
#| warning: false
library(here)
here::i_am("Measurement-issues_SHARE-analysis.Rproj")
```

```{r}
#| message: false
#| warning: false

# Load required libraries
library(dplyr)
library(ggplot2)
library(stringr)
library(scales)
library(ggthemes)
library(hrbrthemes)
library(viridis)
library(RColorBrewer)
library(forcats)
library(gt)
library(broom)
library(ineq)
library(stargazer)

# Import dataset
load("C:/Users/alili/Documents/Etudes/Dauphine/Master Quantitative Economics/S2/4. Measurement issues/Final project/Measurement-issues_SHARE-analysis/easySHARE data/easySHARE_rel9_0_0.rda")
easySHARE <- easySHARE_rel9_0_0

# Transforming -13 codes as NA for income and income percentile
easySHARE <- easySHARE %>%
  mutate(across(starts_with("income_pct_w"), ~ ifelse(. %in% c(-13,-10, -12, -15), NA, .))) %>%
  mutate(thinc_m = ifelse(thinc_m %in% c(-10, -13, -12, -15), NA, thinc_m))%>%
  mutate(hc002_mod = ifelse(hc002_mod %in% c(-10, -13, -12,-15), NA, hc002_mod)) %>%
  mutate(age = ifelse(age %in% c(-10, -13, -12,-15), NA, age))%>%
  mutate(int_month = ifelse(int_month %in% c(-10, -13, -12,-15), NA, int_month))%>%
  mutate(chronic_mod = ifelse(chronic_mod %in% c(-10, -13, -12,-15), NA, chronic_mod))%>%
  mutate(sphus = ifelse(sphus %in% c(-10, -13, -12,-15), NA, sphus))%>%
  mutate(eurod = ifelse(eurod %in% c(-10, -13, -12,-15), NA, eurod))%>%
  mutate(isced1997_r = ifelse(isced1997_r %in% c(-10, -13, -12,-15), NA, isced1997_r))%>%
  mutate(books_age10 = ifelse(books_age10 %in% c(-10, -13, -12,-15), NA, books_age10))%>%
  mutate(adlwa = ifelse(adlwa %in% c(-10, -13, -12,-15), NA, adlwa))%>%
  mutate(smoking = ifelse(smoking %in% c(-10, -13, -12,-15), NA, smoking))


# Subset of Sweden
sweden_data <- subset(easySHARE, substr(mergeid, 1, 2) == "SE") %>%
  mutate(gender = factor(female, levels = c(0, 1), labels = c("Male", "Female")))
```

The country chosen for most of the analytics asked is Sweden.


# I. DESCRIPTIVE STATISTICS

```{r}
#| message: false
#| warning: false

breaks <- seq(40, 100, by = 5)
labels <- paste0(breaks[-length(breaks)], "-", breaks[-1] - 1)

# Step 1: Prep main dataset
income_by_age_gender <- sweden_data %>%
  filter(!is.na(female) & !is.na(age) & !is.na(thinc_m)) %>%
  mutate(
    age_group = cut(age, breaks = breaks, labels = labels),
    gender = factor(female, levels = c(0, 1), labels = c("Male", "Female"))
  ) %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group, gender) %>%
  summarise(mean_income = mean(thinc_m, na.rm = TRUE), .groups = "drop")

# Step 2: Count observations per age group
obs_by_age <- sweden_data %>%
  filter(!is.na(age)) %>%
  mutate(age_group = cut(age, breaks = breaks, labels = labels)) %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group) %>%
  summarise(n_obs = n(), .groups = "drop")

# Step 3: Plot
ggplot() +
  # Mean income bars by gender
  geom_bar(data = income_by_age_gender, 
           aes(x = age_group, y = mean_income, fill = gender), 
           stat = "identity", position = "dodge") +
  # Add line for number of observations (scaled for visibility)
  geom_line(data = obs_by_age, 
            aes(x = age_group, y = n_obs * 10, group = 1), 
            color = "black", size = 1) +
  geom_point(data = obs_by_age, 
             aes(x = age_group, y = n_obs * 10), 
             color = "black", size = 2) +
  scale_y_continuous(
    name = "Mean Household Income",
    sec.axis = sec_axis(~./10, name = "Number of Observations")
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Age Group", fill = "Gender") +
  theme_tufte() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.minor.y = element_blank()
  )
```


```{r setup-values, include=FALSE}
#| message: false
#| warning: false

n_obs_sweden <- sweden_data %>%
  filter(!is.na(thinc_m)) %>%
  nrow()

mean_income_sweden <- sweden_data %>%
  summarise(mean = mean(thinc_m, na.rm = TRUE)) %>%
  pull(mean)

mean_age_sweden <- sweden_data %>%
  summarise(mean = round(mean(age, na.rm = TRUE))) %>%
  pull(mean)

comma(n_obs_sweden)
comma(mean_income_sweden, accuracy = 1)
```

The subset of Sweden data contains `r scales::comma(n_obs_sweden)` observations with an average income of `r scales::comma(mean_income_sweden, accuracy = 1)`€, from a population of `r mean_age_sweden` years on average.

```{r}
#| message: false
#| warning: false

# 1. Define employment labels
employment_labels <- c(
  "1" = "Retired",
  "2" = "Employed",
  "3" = "Unemployed",
  "4" = "Disabled",
  "5" = "Homemaker",
  "97" = "Other"
)

# 2. Preparing info on employement type by age
employment_age <- sweden_data %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = breaks, labels = labels),
    employment = factor(as.character(ep005_), levels = names(employment_labels), labels = employment_labels),
    employment = fct_explicit_na(employment, na_level = "Unknown")
  ) %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group, employment) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(age_group) %>%
  mutate(percent = n / sum(n) * 100)


# 3. Preparing info on median income by age
sweden_data <- sweden_data %>%
  rowwise() %>%
  mutate(avg_income_percentile = mean(c_across(starts_with("income_pct_w")), na.rm = TRUE)) %>%
  ungroup()

median_inc_percentile_by_age <- sweden_data %>%
  filter(!is.na(age) & !is.na(avg_income_percentile)) %>%
  mutate(age_group = cut(age, breaks = breaks, labels = labels)) %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group) %>%
  summarise(median_income_percentile = median(avg_income_percentile, na.rm = TRUE), .groups = "drop")


# 4. Plot
ggplot(employment_age, aes(x = age_group, y = percent, fill = employment)) +
  geom_bar(stat = "identity") +
  geom_line(
    data = median_inc_percentile_by_age,
    aes(x = age_group, y = median_income_percentile * 10, group = 1),
    inherit.aes = FALSE,
    color = "darkblue",
    size = 1
  ) +
  geom_point(
    data = median_inc_percentile_by_age,
    aes(x = age_group, y = median_income_percentile * 10),
    inherit.aes = FALSE,
    color = "darkblue",
    size = 2
  ) +
  scale_y_continuous(
    name = "Percentage (Employment Status)",
    sec.axis = sec_axis(~ . / 10, name = "Median Income percentile")  # Match scaling
  ) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Employment Situation by Age Group and Median Income Percentile",
    x = "Age Group",
    fill = "Employment Status"
  ) +
  theme_tufte() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.minor.y = element_blank()
  )
```


# II. Average household income across 3 countries 

```{r}
#| message: false
#| warning: false

# Mapping the countries of interest
mapping_country <- c(
  "17" = "France",
  "19" = "Greece",
  "13" = "Sweden")
  
# Creating and summarising the subset
income_comparison <- easySHARE %>%
  filter(
    country %in% c("13", "17", "19"),
        wave == 6, 
    !is.na(thinc_m)
  ) %>%
  mutate(country_name = mapping_country[as.character(country)])

income_summary <- income_comparison %>%
  group_by(country_name) %>%
  summarise(
    mean_income = mean(thinc_m, na.rm = TRUE),
    sd_income = sd(thinc_m, na.rm = TRUE),
    n = n(),
    se = sd_income / sqrt(n),  # Standard Error
    .groups = "drop"
  )

# Plotting
ggplot(income_summary, aes(x = country_name, y = mean_income, fill = country_name)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean_income - se, ymax = mean_income + se), width = 0.2) +
  geom_text(aes(label = comma(round(mean_income, 0))), vjust = -0.75, size = 2.5) +
  labs(
    title = "Mean Household Income by Country (Wave 6)",
    x = "Country",
    y = "Mean Household Income (€)"
  ) +
  scale_fill_brewer(palette = "Pastel2")+
  theme_tufte() +
  theme(legend.position = "none")

```
```{r}
#| message: false
#| warning: false

# Anova test
anova_result <- aov(thinc_m ~ country_name, data = income_comparison)

# Tukey test
tukey_result <- TukeyHSD(anova_result)
tukey_df <- as.data.frame(tukey_result$country_name)
tukey_df$comparison <- rownames(tukey_df)
rownames(tukey_df) <- NULL
colnames(tukey_df)[colnames(tukey_df) == "p adj"] <- "p_value"

# Table Tukey
gt(tukey_df) %>%
  tab_header(
    title = "Tukey HSD Results"
  ) %>%
  fmt_number(columns = c(diff, lwr, upr, p_value), decimals = 3)

```
Sweden-France not statistically different
But the 2 other pairs are
The 3rd place of the ranking would for sure be the Greece but Sweden and France are not rankable, apart from being ex-aeco.


# III. OLS regression explaining the frequency of visits to the doctor

```{r, results='asis'}
#| message: false
#| warning: false

OLS <- lm(hc002_mod ~ age + int_month + chronic_mod + sphus + eurod + isced1997_r + books_age10 + adlwa + smoking, data = sweden_data)

# female + ch001_ + childhood_health + bmi2 + br015_ + ep005_

#tidy(OLS)

stargazer(OLS,
         type = "latex",
         title = "Regression Results")
```



hc002_mod contains the number of doctor visits within the past year

gender + nb of children + age + interview month + education + childhood health status + self-perceived health + body mass index + vigorous sports + Current job situation 



# IV. Effect of a subsidy on inequality - poverty

## a. Effect of a subsidy on income

```{r}
#| message: false
#| warning: false

# create new variable of subsidy income
sweden_data <- sweden_data %>%
  mutate(
    subsidy_income = hc002_mod * 10,
    income_subsidized = thinc_m + subsidy_income
  )
mean_increase <- round(mean(sweden_data$subsidy_income, na.rm = TRUE),0)
```

In the whole population, the implementation of such a susidy increases their income by `r mean_increase`€.

```{r}
#| message: false
#| warning: false

# Analysis by age group
income_by_age <- sweden_data %>%
  filter(!is.na(age), !is.na(subsidy_income), !is.na(gender)) %>%
  mutate(age_group = cut(age, breaks = breaks, labels = labels)) %>%
  group_by(age_group, gender) %>%
  summarise(avg_gain = mean(subsidy_income, na.rm = TRUE))

ggplot(na.omit(income_by_age), aes(x = age_group, y = avg_gain, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Subsidy by Age and Gender",
    x = "Age Group",
    y = "Average Subsidy gain (€)",
    fill = "Gender"
  ) +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
#| message: false
#| warning: false

# Analysis by income percentile
sweden_data <- sweden_data %>%
  filter(!is.na(avg_income_percentile), !is.na(gender)) %>%
  mutate(income_percentile_group = cut(
    avg_income_percentile,
    breaks = seq(1, 10, by = 1),
    include.lowest = TRUE,
    right = FALSE,
    labels = paste0("Decile ", 1:9)
  ))

# Summarising by group
subsidy_by_decile_gender <- sweden_data %>%
  group_by(income_percentile_group, gender) %>%
  summarise(
    avg_subsidy = mean(subsidy_income, na.rm = TRUE),
    avg_income_before = mean(thinc_m, na.rm = TRUE),
    avg_income_after = mean(income_subsidized, na.rm = TRUE),
    .groups = "drop"
  )

# Plotting
ggplot(subsidy_by_decile_gender, aes(x = income_percentile_group, y = avg_subsidy, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Subsidy by Income Decile and Gender",
    x = "Income Decile Group",
    y = "Average Subsidy (€)",
    fill = "Gender"
  ) +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## b. Effect of the subsidy on poverty

“At risk of poverty” = income below 60% of median equivalised income
compute poverty rate before and after the subsidy using this approach.

```{r}
#| message: false
#| warning: false

# Defining the poverty lines
poverty_line_before <- 0.6 * median(sweden_data$thinc_m, na.rm = TRUE)
poverty_line_after <- 0.6 * median(sweden_data$income_subsidized, na.rm = TRUE)

# Computing poverty rate before and after
sweden_data <- sweden_data %>%
  mutate(
    is_poor_before = thinc_m < poverty_line_before,
    is_poor_after = income_subsidized < poverty_line_after
  )

poverty_rate_before <- mean(sweden_data$is_poor_before, na.rm = TRUE) * 100
poverty_rate_after <- mean(sweden_data$is_poor_after, na.rm = TRUE) * 100
poverty_change <- poverty_rate_before - poverty_rate_after
```
Using the EU's relative poverty threshold (60% of the median income), the poverty rate in Sweden is estimated at `r round(poverty_rate_before, 2)`% before the subsidy. After implementing a €10 per doctor visit subsidy, the poverty rate falls to `r round(poverty_rate_after, 2)`%, indicating a `r round(poverty_change, 2)` percentage point reduction. This result suggests that the subsidy has a modest but measurable effect on poverty alleviation, especially among those with more frequent health needs.


## c. Redistributive effect of a subsidy

Gini coeff : 
sensitive to the middle of the distribution, and responds clearly to redistributive policies like your €10-per-visit subsidy
--> compute the coef before and after the subsidy

```{r gini-calc, include=FALSE}
#| message: false
#| warning: false

gini_before <- ineq::Gini(na.omit(sweden_data$thinc_m)) * 100
gini_after <- ineq::Gini(na.omit(sweden_data$income_subsidized)) * 100
gini_change <- gini_before - gini_after
```

The Gini coefficient **before** the subsidy is `r round(gini_before, 2)`%, and **after** the subsidy it is `r round(gini_after, 2)`%.  
This corresponds to a **reduction in inequality of** `r round(gini_change, 2)` percentage points.


```{r}
#| message: false
#| warning: false

plot(Lc(sweden_data$thinc_m), col = "red", lwd = 2, main = "Lorenz Curve: Before vs After Subsidy")
lines(Lc(sweden_data$income_subsidized), col = "blue", lwd = 2)
legend("topleft", legend = c("Before", "After"), col = c("red", "blue"), lwd = 2)
```
The change of inequality is so small that the curves don't even show any, they are confounded in one single curve. If there was a significant change, the blue should be closer to the black line, showing perfect equality.


# V. Behavioral response to the subsidy

## a. Methodology and simulation of behavioral response

## b. Subsidy progressive or regressive ?





